---
title: "Final project"
author: "Katherine Mercurio"
date: "May 8, 2017"
output:
  pdf_document:
    toc: yes
  html_document: default
---
Introduction

I decided to focus on the demographic characteristics and locations of juveniles who 
committed arson between 2012 and 2015. I chose these years because they are the most 
recent years, so they would be the most useful in predicting the rates of arson in 
2016, if someone wanted to use this data to figure that out.
My specific question was which race, gender, and state have the highest rates of arson. I found out that white males and Iowa have the higest rates for people and states respectively. 

Tidying Data and Creating Preliminary Tables 

I spent a lot of time tidying my data because I wanted to make sure the data would 
convey the characteristics I chose to focus on. I began by changing the number of 
digits R requires to use scientific notation so that all of the money data would be 
in standard notation. Then, I began reading in the arson, basic incident, FD header,
and juvenile files. I kept them separated by year. Then, I began joining the tables 
together by attribute. I had some trouble with the 2012 version column, so I deleted
that column from each table. 
```{r code = readLines('initialtables.R'), message=FALSE, warning=FALSE, include=FALSE, error = TRUE}
ar2012 <- read.table('fire2012/arson.txt', sep = '^', header = TRUE, stringsAsFactors = FALSE)
ar2013 <- read.table('fire2013/arson.txt', sep = '^', header = TRUE, stringsAsFactors = FALSE)
ar2014 <- read.table('fire2014/arson.txt', sep = '^', header = TRUE, stringsAsFactors = FALSE)
ar2015 <- read.table('fire2015/arson.txt', sep = '^', header = TRUE, stringsAsFactors = FALSE)


bi2012 <- read.table('fire2012/basicincident.txt', sep = '^', header = TRUE, stringsAsFactors = FALSE)
bi2013 <- read.table('fire2013/basicincident.txt', sep = '^', header = TRUE, stringsAsFactors = FALSE)
bi2014 <- read.table('fire2014/basicincident.txt', sep = '^', header = TRUE, stringsAsFactors = FALSE)
bi2015 <- read.table('fire2015/basicincident.txt', sep = '^', header = TRUE, stringsAsFactors = FALSE)


aj2012 <- read.table('fire2012/arsonjuvsub.txt', sep = '^', header = TRUE, stringsAsFactors = FALSE)
aj2013 <- read.table('fire2013/arsonjuvsub.txt', sep = '^', header = TRUE, stringsAsFactors = FALSE)
aj2014 <- read.table('fire2014/arsonjuvsub.txt', sep = '^', header = TRUE, stringsAsFactors = FALSE)
aj2015 <- read.table('fire2015/arsonjuvsub.txt', sep = '^', header = TRUE, stringsAsFactors = FALSE)


header2012 <- read.table('fire2012/fdheader.txt', sep = '^', header = TRUE, stringsAsFactors = FALSE)
header2013 <- read.table('fire2013/fdheader.txt', sep = '^', header = TRUE, stringsAsFactors = FALSE)
header2014 <- read.table('fire2014/fdheader.txt', sep = '^', header = TRUE, stringsAsFactors = FALSE)
header2015 <- read.table('fire2015/fdheader.txt', sep = '^', header = TRUE, stringsAsFactors = FALSE)


# joining basic incident and header
bih12 <- inner_join(bi2012, header2012)
bih12$VERSION <- NULL
bih13 <- inner_join(bi2013, header2013)
bih14 <- inner_join(bi2014, header2014)
bih15 <- inner_join(bi2015, header2015)


# joining basic incident, header, and arson
ars2012 <- inner_join(bih12, ar2012)
ars2012$VERSION <- NULL
ars2013 <- inner_join(bih13, ar2013)
ars2014 <- inner_join(bih14, ar2014)
ars2015 <- inner_join(bih15, ar2015)


# joining basic incident, header, arson, and juv
arson2012 <- inner_join(ars2012, aj2012)
arson2013 <- inner_join(ars2013, aj2013)
arson2014 <- inner_join(ars2014, aj2014)
arson2015 <- inner_join(ars2015, aj2015)

```

Once the tables each were how I wanted them by year, I wanted to join them together. 
But, since 2012-2015 is a lot of data, and I only want to focus on juveniles anyway, 
I made a subset of each table first to ensure I only had children 18 and younger. 
Then, I fixed the dates using lubridate to be split into different columns because 
I wanted to analyze the rates of arson during different months/seasons. Once my data 
was subset-ted and the dates were fixed, I joined the tables into one 2012-2015 data 
table called full. 
```{r code = readLines('initialtables.R'), error = TRUE}
under1812 <- subset(x = arson2012, subset = arson2012$AGE <= 18)
under1813 <- subset(x = arson2013, subset = arson2013$AGE <= 18)
under1814 <- subset(x = arson2014, subset = arson2014$AGE <= 18)
under1815 <- subset(x = arson2015, subset = arson2015$AGE <= 18)

# now that i have 3 tables with all cases of arson for these years 
# commited by youth 18 or under,
# I'm going to split up the dates and begin anaylzing the data

under1812$INC_DATE <- mdy(under1812$INC_DATE)
under1813$INC_DATE <- mdy(under1813$INC_DATE)
under1814$INC_DATE <- mdy(under1814$INC_DATE)
under1815$INC_DATE <- mdy(under1815$INC_DATE)


under1812 <- under1812 %>% separate(col = INC_DATE, c("Year", "Month", "Date"))
under1813 <- under1813 %>% separate(col = INC_DATE, c("Year", "Month", "Date"))
under1814 <- under1814 %>% separate(col = INC_DATE, c("Year", "Month", "Date"))
under1815 <- under1815 %>% separate(col = INC_DATE, c("Year", "Month", "Date"))
under1813$VERSION <- NULL
under1814$VERSION <- NULL
under1815$VERSION <- NULL
under1812$VERSION <- NULL

full <- rbind(under1812, under1813, under1814, under1815)
full <- as.data.table(full)
```

Once I had my full data table, I started determining which columns would not be of interest. 
I did not do this right away; I read through some of the PDFs that came with the data, and 
then I looked at different codes and used those to determine which columns I would need. 
I added more columns to this section of code as I continued to narrow my focus. 

```{r code = readLines('initialtables.R'), message=FALSE, warning=FALSE, include=FALSE, error = TRUE}
# deleting columns I will not use in analysis
full$ALARM <- NULL
full$ARRIVAL <- NULL
full$FD_STR_NO <- NULL
full$FD_STR_PRE <- NULL
full$FD_STREET <- NULL
full$FD_STR_TYP <- NULL
full$ADD_WILD <- NULL
full$INC_CONT <- NULL
full$LU_CLEAR <- NULL
full$SHIFT <- NULL
full$ALARMS <- NULL
full$FD_PHONE <- NULL
full$FD_EMAIL <- NULL
full$CENSUS <- NULL
full$FD_STR_SUF <- NULL
full$FD_FAX <- NULL
full$LAB_USED1 <- NULL
full$LAB_USED2 <- NULL
full$LAB_USED3 <- NULL
full$LAB_USED4 <- NULL
full$LAB_USED5 <- NULL
full$LAB_USED6 <- NULL
full$EXP_NO <- NULL
full$DISTRICT <- NULL
full$INIT_OB1 <- NULL
full$INIT_OB2 <- NULL
full$INIT_OB3 <- NULL
full$INIT_OB4 <- NULL
full$INIT_OB5 <- NULL
full$INIT_OB6 <- NULL
full$INIT_OB7 <- NULL
full$INIT_OB8 <- NULL
full$ACT_TAK1 <- NULL
full$ACT_TAK2 <- NULL
full$ACT_TAK3 <- NULL
full$NO_STATION <- NULL
full$ENTRY_METH<- NULL
full$PROP_OWNER<- NULL
full$AID<- NULL
full$INC_TYPE<- NULL
full$INV_INFO1 <- NULL
full$INV_INFO2 <- NULL
full$INV_INFO3 <- NULL
full$INV_INFO4 <- NULL
full$INV_INFO5 <- NULL
full$INV_INFO6 <- NULL
full$INV_INFO7 <- NULL
full$INV_INFO8 <- NULL
full$GRP_INVOL1 <- NULL
full$GRP_INVOL2 <- NULL
full$GRP_INVOL3 <- NULL
full$JUV_DISPO <- NULL
full$DEVI_CONT <- NULL
full$DEVI_FUEL <- NULL
full$DEVI_IGNIT <- NULL
full$EXT_FIRE <- NULL
full$RISK_FACT8 <- NULL
full$FF_DEATH <- NULL
full$CASE_STAT <- NULL
full$APP_MOD <- NULL
full$DEPT_STA <- NULL
full$AVAIL_MFI <- NULL
full$SUP_APP <- NULL
full$EMS_APP <- NULL
full$OTH_APP <- NULL
full$SUP_PER <- NULL
full$EMS_PER <- NULL
full$OTH_PER <- NULL
full$RESOU_AID <- NULL
full$OTH_DEATH <- NULL
full$FF_INJ <- NULL
full$OTH_INJ <- NULL
full$NO_VOL_PDC <- NULL
full$DET_ALERT <- NULL
full$HAZ_REL <- NULL
full$MIXED_USE <- NULL
full$FD_FIP_CTY <- NULL
full$NO_PD_FF <- NULL
full$NO_VOL_FF <- NULL
full$PROP_VAL <- NULL
full$CONT_VAL <- NULL
full$PROP_USE <- NULL
```

I continued tidying my data by changing the values shown in the data table so that you can see exactly what the code represents. I also changed some column names based on the coded values the columns represented. I decided to make most NA data "unknown". The comments clearly note which factor is being changed. 

```{r code = readLines('initialtables.R'), message=FALSE, warning=FALSE, include=FALSE, error = TRUE}
# changing codes to show their definition 

factor(full$MOT_FACTS1)
factor(full$MOT_FACTS2)
factor(full$MOT_FACTS3)
factor(full$RISK_FACT1)
factor(full$RISK_FACT2)
factor(full$RISK_FACT3)
factor(full$RISK_FACT4)
factor(full$RISK_FACT5)
factor(full$RISK_FACT6)
factor(full$RISK_FACT7)
factor(full$RISK_FACT8)

# RISK FACTORS
levels(factor(full$RISK_FACT1))
levels(full$RISK_FACT1) <- c(levels(full$RISK_FACT1), "mild", "moderate", "extreme", 'unknown')
full$RISK_FACT1[full$RISK_FACT1 == "1 -- Mild curiosity about fire"] <- "mild"
full$RISK_FACT1[full$RISK_FACT1 == "2 -- Moderate curiosity about fire"] <- "moderate"
full$RISK_FACT1[full$RISK_FACT1 == "3 -- Extreme curiosity about fire"] <- "extreme"
full$RISK_FACT1[full$RISK_FACT1 == "U -- Unknown"] <- 'unknown'
full$RISK_FACT1[full$RISK_FACT1 == "Z"] <- 'unknown'
full$RISK_FACT1[full$RISK_FACT1 == ''] <- 'unknown'
full$RISK_FACT1[full$RISK_FACT1 == 'unknown'] <- NA
full$RISK_FACT1[full$RISK_FACT1 == "1"] <- "mild"
full$RISK_FACT1[full$RISK_FACT1 == "2"] <- "moderate"
full$RISK_FACT1[full$RISK_FACT1 == "3"] <- "extreme"
colnames(full)[which(names(full) == "RISK_FACT1")] <- "Fire Curiosity Level"
full$`Fire Curiosity Level`[is.na(full$`Fire Curiosity Level`)] <- 'unknown'
#ADD/ADHD
colnames(full)[which(names(full) == "RISK_FACT2")] <- "ADD_ADHD"
full$ADD_ADHD[full$ADD_ADHD == '4'] <- 'yes'
full$ADD_ADHD[full$ADD_ADHD != 'yes'] <- 'no'
full$ADD_ADHD[is.na(full$ADD_ADHD)] <- 'no'

# out of school trouble
colnames(full)[which(names(full) == "RISK_FACT3")] <- "TroubleOutsideSchool"
full$TroubleOutsideSchool[full$TroubleOutsideSchool == '5'] <- 'yes'
full$TroubleOutsideSchool[full$TroubleOutsideSchool != 'yes'] <- 'no'
full$TroubleOutsideSchool[is.na(full$TroubleOutsideSchool)] <- 'no'


# history of stealing or shoplifting 
colnames(full)[which(names(full) == "RISK_FACT4")] <- "StealingShoplifting"
full$StealingShoplifting[full$StealingShoplifting == '6'] <- 'yes'
full$StealingShoplifting[full$StealingShoplifting != 'yes'] <- 'no'
full$StealingShoplifting[is.na(full$StealingShoplifting)] <- 'no'


#history of physical assault
colnames(full)[which(names(full) == "RISK_FACT5")] <- "PhysicalAssault"
full$PhysicalAssault[full$PhysicalAssault == '7'] <- 'yes'
full$PhysicalAssault[full$PhysicalAssault != 'yes'] <- 'no'
full$PhysicalAssault[is.na(full$PhysicalAssault)] <- 'no'

# history of fireplay/firesetting
colnames(full)[which(names(full) == "RISK_FACT6")] <- "Firesetting"
full$Firesetting[full$Firesetting == '8'] <- 'yes'
full$Firesetting[full$Firesetting != 'yes'] <- 'no'
full$Firesetting[is.na(full$Firesetting)] <- 'no'


# transiency
colnames(full)[which(names(full) == "RISK_FACT7")] <- "Transiency"
full$Transiency[full$Transiency == '9'] <- 'yes'
full$Transiency[full$Transiency != 'yes'] <- 'no'
full$Transiency[is.na(full$Transiency)] <- 'no'


# MOTIVATION FACTORS 1
levels(full$MOT_FACTS1)
levels(full$MOT_FACTS1) <- c(levels(full$MOT_FACTS1), 'unknown', 'extortion', 'labor unrest', 'insurance fraud', 
                                 'intimidation', 'void contract/lease', 'foreclosed property', 'personal', 'hate crime', 
                                 'institutional', 'societal', 'protest', 'civil unrest', 'fireplay/curiosity', 
                                 'vanity/recognition', 'thrills', 'attention/sympathy', 'homicide', 'suicide', 
                                 'domestic violence', 'burglary', 'homicide concealment', 'burglary concealment',
                                 'auto theft concealment', 'destroy records/evidence', 'other')
full$MOT_FACTS1[full$MOT_FACTS1 == ''] <- 'unknown'
full$MOT_FACTS1[full$MOT_FACTS1 == 'UU'] <- 'unknown'
full$MOT_FACTS1[full$MOT_FACTS1 == '11'] <- 'extortion'
full$MOT_FACTS1[full$MOT_FACTS1 == '12'] <- 'labor unrest'
full$MOT_FACTS1[full$MOT_FACTS1 == '13'] <- 'insurance fraud'
full$MOT_FACTS1[full$MOT_FACTS1 == '14'] <- 'intimidation'
full$MOT_FACTS1[full$MOT_FACTS1 == '15'] <- 'void contract/lease'
full$MOT_FACTS1[full$MOT_FACTS1 == '16'] <- 'foreclosed property'
full$MOT_FACTS1[full$MOT_FACTS1 == '21'] <- 'personal'
full$MOT_FACTS1[full$MOT_FACTS1 == '22'] <- 'hate crime'
full$MOT_FACTS1[full$MOT_FACTS1 == '23'] <- 'institutional'
full$MOT_FACTS1[full$MOT_FACTS1 == '24'] <- 'societal'
full$MOT_FACTS1[full$MOT_FACTS1 == '31'] <- 'protest'
full$MOT_FACTS1[full$MOT_FACTS1 == '32'] <- 'civil unrest'
full$MOT_FACTS1[full$MOT_FACTS1 == '41'] <- 'fireplay/curiosity'
full$MOT_FACTS1[full$MOT_FACTS1 == '42'] <- 'vanity/recognition'
full$MOT_FACTS1[full$MOT_FACTS1 == '43'] <- 'thrills'
full$MOT_FACTS1[full$MOT_FACTS1 == '44'] <- 'attention/sympathy'
full$MOT_FACTS1[full$MOT_FACTS1 == '51'] <- 'homicide'
full$MOT_FACTS1[full$MOT_FACTS1 == '52'] <- 'suicide'
full$MOT_FACTS1[full$MOT_FACTS1 == '53'] <- 'domestic violence'
full$MOT_FACTS1[full$MOT_FACTS1 == '54'] <- 'burglary'
full$MOT_FACTS1[full$MOT_FACTS1 == '61'] <- 'homicide concealment'
full$MOT_FACTS1[full$MOT_FACTS1 == '62'] <- 'burglary concealment'
full$MOT_FACTS1[full$MOT_FACTS1 == '63'] <- 'auto theft concealment'
full$MOT_FACTS1[full$MOT_FACTS1 == '64'] <- 'destroy records/evidence'
full$MOT_FACTS1[full$MOT_FACTS1 == '00'] <- 'other'
full$MOT_FACTS1[is.na(full$MOT_FACTS1)] <- 'unknown'


# MOTIVATION FACTORS 2

levels(full$MOT_FACTS2)
levels(full$MOT_FACTS2) <- c(levels(full$MOT_FACTS2),'labor unrest', 'insurance fraud', 'intimidation', 'void contract/lease',
                                'personal', 'institutional', 'societal', 'protest', 
                                'fireplay/curiosity', 'vanity/recognition', 'thrills', 'attention/sympathy', 
                                'homicide', 'suicide', 'domestic violence', 'burglary', 'burglary concrealment',
                                'auto theft concealment', 'destroy records/evidence')

full$MOT_FACTS2[full$MOT_FACTS2 == '12'] <- 'labor unrest'
full$MOT_FACTS2[full$MOT_FACTS2 == '13'] <- 'insurance fraud'
full$MOT_FACTS2[full$MOT_FACTS2 == '14'] <- 'intimidation'
full$MOT_FACTS2[full$MOT_FACTS2 == '15'] <- 'void contract/lease'
full$MOT_FACTS2[full$MOT_FACTS2 == '21'] <- 'personal'
full$MOT_FACTS2[full$MOT_FACTS2 == '23'] <- 'institutional'
full$MOT_FACTS2[full$MOT_FACTS2 == '24'] <- 'societal'
full$MOT_FACTS2[full$MOT_FACTS2 == '31'] <- 'protest'
full$MOT_FACTS2[full$MOT_FACTS2 == '41'] <- 'fireplay/curiosity'
full$MOT_FACTS2[full$MOT_FACTS2 == '42'] <- 'vanity/recognition'
full$MOT_FACTS2[full$MOT_FACTS2 == '43'] <- 'thrills'
full$MOT_FACTS2[full$MOT_FACTS2 == '44'] <- 'attention/sympathy'
full$MOT_FACTS2[full$MOT_FACTS2 == '51'] <- 'homicide'
full$MOT_FACTS2[full$MOT_FACTS2 == '52'] <- 'suicide'
full$MOT_FACTS2[full$MOT_FACTS2 == '53'] <- 'domestic violence'
full$MOT_FACTS2[full$MOT_FACTS2 == '54'] <- 'burglary'
full$MOT_FACTS2[full$MOT_FACTS2 == '62'] <- 'burglary concealment'
full$MOT_FACTS2[full$MOT_FACTS2 == '63'] <- 'auto theft concealment'
full$MOT_FACTS2[full$MOT_FACTS2 == '64'] <- 'destroy records/evidence'
full$MOT_FACTS2[is.na(full$MOT_FACTS2)] <- 'unknown'
full$MOT_FACTS2[full$MOT_FACTS2 == ''] <- 'unknown'


# MOTIVATION FACTORS 3

levels(factor(full$MOT_FACTS3))
levels(full$MOT_FACTS3) <- c(levels(full$MOT_FACTS3), 'insurance fraud', 'personal', 'fireplay/curiosity', 
                                 'vanity/recognition', 'thrills', 'attention/sympathy', 'sexual excitement', 'suicide', 
                                 'domestic violence', 'auto theft concealment')

full$MOT_FACTS3[full$MOT_FACTS3 == '13'] <- 'insurance fraud'
full$MOT_FACTS3[full$MOT_FACTS3 == '21'] <- 'personal'
full$MOT_FACTS3[full$MOT_FACTS3 == '41'] <- 'fireplay/curiosity'
full$MOT_FACTS3[full$MOT_FACTS3 == '42'] <- 'vanity/recognition'
full$MOT_FACTS3[full$MOT_FACTS3 == '43'] <- 'thrills'
full$MOT_FACTS3[full$MOT_FACTS3 == '44'] <- 'attention/sympathy'
full$MOT_FACTS3[full$MOT_FACTS3 == '45'] <- 'sexual excitement'
full$MOT_FACTS3[full$MOT_FACTS3 == '52'] <- 'suicide'
full$MOT_FACTS3[full$MOT_FACTS3 == '53'] <- 'domestic violence'
full$MOT_FACTS3[full$MOT_FACTS3 == '63'] <- 'auto theft concealment'
full$MOT_FACTS3[full$MOT_FACTS3 == '14'] <- 'intimidation'
full$MOT_FACTS3[full$MOT_FACTS3 == '32'] <- 'civil unrest'
full$MOT_FACTS3[full$MOT_FACTS3 == '64'] <- 'destroy records/evidence'
full$MOT_FACTS3[full$MOT_FACTS3 == '51'] <- 'homicide'
full$MOT_FACTS3[full$MOT_FACTS3 == '23'] <- 'institutional'
full$MOT_FACTS3[full$MOT_FACTS3 == '54'] <- 'burglary'
full$MOT_FACTS3[full$MOT_FACTS3 == ''] <- 'unknown'
full$MOT_FACTS3[is.na(full$MOT_FACTS3)] <- 'unknown'


# FAMILY TYPE

levels(factor(full$FAM_TYPE))
levels(full$FAM_TYPE) <- c(levels(full$FAM_TYPE), 'other', 'single-parent family', 'foster parents', 'two-parent family',
                               'extended family', 'no family unit', 'unknown')
full$FAM_TYPE[full$FAM_TYPE == '0'] <- "other"
full$FAM_TYPE[full$FAM_TYPE == '1'] <- "single-parent family"
full$FAM_TYPE[full$FAM_TYPE == '2'] <- "foster parents"
full$FAM_TYPE[full$FAM_TYPE == '3'] <- "two-parent family"
full$FAM_TYPE[full$FAM_TYPE == '4'] <- "extended family"
full$FAM_TYPE[full$FAM_TYPE == 'N'] <- "no family unit"
full$FAM_TYPE[full$FAM_TYPE == 'U'] <- "unknown"
full$FAM_TYPE[full$FAM_TYPE == ''] <- "unknown"
full$FAM_TYPE[full$FAM_TYPE == 'unknown'] <- NA
full$FAM_TYPE[is.na(full$FAM_TYPE)] <- 'unknown'


# RACE

levels(factor(full$RACE))
levels(full$RACE) <- c(levels(full$RACE), 'other/multiracial', 'White', 'Black or African American', 'Asian', 
                           'Native Hawaiian/Pacific Islander', "American Indian or Alaska Native", "undetermined")
full$RACE[full$RACE == '0'] <- 'other/multiracial'
full$RACE[full$RACE == '1'] <- 'White'
full$RACE[full$RACE == '2'] <- 'Black or African American'
full$RACE[full$RACE == '3'] <-"American Indian or Alaska Native"
full$RACE[full$RACE == '4'] <- "Asian"
full$RACE[full$RACE == '5'] <- 'Native Hawaiian/Pacific Islander'
full$RACE[full$RACE == 'U'] <- "undetermined"
full$RACE[is.na(full$RACE)] <- 'undetermined'
full$RACE[full$RACE == ''] <-  'undetermined'


# Ethnicity

levels(factor(full$ETHNICITY))
levels(full$ETHNICITY) <- c(levels(full$ETHNICITY), 'Hispanic/Latino', "Non Hispanic/Latino", 'unknown')
full$ETHNICITY[full$ETHNICITY == '0'] <- 'Non Hispanic/Latino'
full$ETHNICITY[full$ETHNICITY == '1'] <- 'Hispanic/Latino'
full$ETHNICITY[full$ETHNICITY == ''] <- 'unknown'
full$ETHNICITY[is.na(full$ETHNICITY)] <- 'unknown'


# GENDER

levels(factor(full$GENDER))
levels(full$GENDER) <- c(levels(full$GENDER), 'Male', 'Female', 'unknown')
full$GENDER[full$GENDER == '1'] <- 'Male'
full$GENDER[full$GENDER == '2'] <- 'Female'
full$GENDER[is.na(full$GENDER)] <- 'unknown'
```

The data set is now adequately cleaned. I wasn't sure whether ages of 0 were typos or babies committing arson, but I left them as is. 

Descriptive Statistics and Analysis 

I decided to make a new data table to use in conjunction with some plots. I grouped by state and played around with different table spreads. I ended up picking 3 states--MA, CT, and NY-- and made a bar graph showing their amounts of arson over the 2012-2015. I was very surprised to find out how high Massachusetts rates of juvenile arson are. I'm not sure if this was an error, or if MA really has that many. I had to load and unload plyr to make the n() function in dplyr work, since plyr overrides the summarise function in dplyr. This first chunk is my new table showing rates per year in each state. 
```{r code = readLines('initialtables.R'), echo=FALSE, error = TRUE}

countingfullN <- full %>% 
  group_by(Year, STATE) %>%
  summarise(number = n())
max(countingfullN$number)



countingfull <- spread(countingfullN, key = Year, value = number)
countingfull[is.na(countingfull)] <- 0
head(countingfull, n = 10)
countingfu <- countingfull[, 2:5]
countingfullT <- cbind(countingfull, Total = rowSums(countingfu))
head(countingfullT, n = 10)
tail(countingfullT, n = 10)
```

This next chunk of code creates the MA, NY, CT plot. 

```{r code = readLines('initialtables.R'), echo=FALSE, error= TRUE}

countingfullsub <- subset(countingfullN, STATE == 'MA')
countingfullsub2 <- subset(countingfullN, STATE == 'NY')
countingfullsub3 <- subset(countingfullN, STATE == "CT")
manyct <- rbind(countingfullsub, countingfullsub2, countingfullsub3)


manyctPLOT <- ggplot(manyct, aes(x=STATE,y = number, fill = Year)) +geom_bar(stat = "identity")
plot(manyctPLOT)

```


```{r code = readLines('initialtables.R'), echo=FALSE, error = TRUE}

racegen <- full %>% 
  ungroup(STATE) %>%
  group_by(RACE, GENDER, Year) %>%
  summarise(number = n()) 
racegen <- subset(racegen, GENDER != 'unknown')
racegen <- subset(racegen, RACE != 'undetermined')
head(racegen, n = 10)
```
I created another table and another plot to summarize the race and gender data for 2012-2015. I used the arson rates as the y axis and the gender as the x axis. I deleted undetermined and unknown races and genders for this plot, because it doesn't really provide any insight into arson rates based on race/gender if they are missing. As I expected, men have much higher rates of arson than women, and white men have the highest rates of arson for all juveniles. 
Here is the graph based on the above data table. 

```{r code = readLines('initialtables.R'), echo=FALSE, error = TRUE}

rgplot <- ggplot(racegen, aes(x = GENDER, y = number, fill = RACE)) +
         geom_bar(stat= 'identity') 

plot(rgplot)

```

Based on my data tables and plots, I calculated some statistics about the data to determine where juvenile arson is most common and who is most likely to commit it. I originally created a table, but I couldn't get it to look how I wanted, so here are the rates of arson as individual vectors. 
```{r code = readLines('initialtables.R'), echo=TRUE, error = TRUE}
#stats 
#2012 
c(countingfullT$STATE[countingfullT$`2012` == max(countingfullT$`2012`)], 
  max(countingfullT$`2012`))
c(countingfullT$STATE[countingfullT$`2012` == min(countingfullT$`2012`)],
  min(countingfullT$`2012`))

#2013
c(countingfullT$STATE[countingfullT$`2013` == max(countingfullT$`2013`)],
  max(countingfullT$`2013`))
c(countingfullT$STATE[countingfullT$`2013` == min(countingfullT$`2013`)],
  min(countingfullT$`2013`))

#2014
c(countingfullT$STATE[countingfullT$`2014` == max(countingfullT$`2014`)],
  max(countingfullT$`2014`))
c(countingfullT$STATE[countingfullT$`2014` == min(countingfullT$`2014`)],
  min(countingfullT$`2014`))

#2015
c(countingfullT$STATE[countingfullT$`2015` == max(countingfullT$`2015`)],
  max(countingfullT$`2015`))
c(countingfullT$STATE[countingfullT$`2015` == min(countingfullT$`2015`)],
  min(countingfullT$`2015`))

# total
c(countingfullT$STATE[countingfullT$Total == max(countingfullT$Total)],
  max(countingfullT$Total))
c(countingfullT$STATE[countingfullT$Total == min(countingfullT$Total)],
  min(countingfullT$Total))
```
Then, I made a map to show using shades of blue which states have the higest rate of arson for each year. 
```{r, echo=FALSE, error=TRUE}
#map data

state <- full[, STATE]
citystate <- as.data.frame(state)
citystate <- unique(citystate)

geocodes <- geocode(as.character(citystate$state))
latlong <- cbind(citystate, geocodes)
fullat <- inner_join(x = countingfullN, y = latlong, by = c("STATE" = "state"))
mapone <- subset(fullat, Year == 2012)
maptwo <- subset(fullat, Year == 2013)
mapthr <- subset(fullat, Year == 2014)
mapfou <- subset(fullat, Year == 2015)
MAmap <- get_map(location= "USA",
                 source= "google", 
                 zoom= 4, scale="auto")

mmap <- ggmap(MAmap)

map1 <- 
  mmap +
  geom_point(data=mapone, aes(x=lon, y=lat, color=number))

map2 <- 
  mmap +
  geom_point(data=maptwo, aes(x=lon, y=lat, color=number))

map3 <- 
  mmap +
  geom_point(data=mapthr, aes(x=lon, y=lat, color=number))

map4 <- 
  mmap +
  geom_point(data=mapfou, aes(x=lon, y=lat, color=number))

plot(map1)
plot(map2) 
plot(map3)
plot(map4)
```


Conclusion:

White males under 18 are the most likely juvenile group to commit arson based on 2012-2015 data. Iowa has the highest rates of juvenile arson over the 2012-2015 period. Take a look at my shiny app to see more ways to organize/look at the data in an interactive data table. 
